---
title: "ROBITT_template"
author: "Rob Boyd"
date: "18 May 2022"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1.	Iteration

### 1.1 ROBITT iteration number

|Iteration |Comments  |
--- | --- | ---|
|1||
	
# 2.	Research statement and pre-bias assessments

## Statistical population of interest

**2.1 Define the statistical target population about which you intend to make inferences.**

|Domain |Extent  | Resolution|
--- | --- | ---|
|Geographic|United Kingdom (UK)|1km grid cells|
|Temporal|1970-2020|Annual increments|
|Taxonomic (or other relevant organismal domain such as functional group etc.)|All species of soldierfly|Species|
|Environmental|Environmental space in the UK|1km to match the geographic resolution|

# Inferential goals

**2.2 What are your inferential goals?** 

To estimate temporal trends in species' occupancy (proportion of occupied grid cells). The individual species trends will be "averaged" to construct a multispecies indicator of change. 

## Data provenance

**2.3 From where were your data acquired (please provide citations, including a DOI, wherever possible)? What are their key features in respect of the inferential aims of your study (see the guidance document for examples)?**

The data are presence-only records of soldierfly occurrences recorded in the UK from 1990-2020.

```{r}
dat <- read.csv("W:/PYWELL_SHARED/Pywell Projects/BRC/_BRC_dataflow/Research Datasets/Soldierflies/2022/Research dataset/Soldierflies_2022.csv")

names(dat)

dat$raw_dataset_name[1]

dat$citation_req[1]

dat$date_of_capture[1]
```

## Data processing

**2.4 Provide details of, and the justification for, all of the steps that you have taken to clean the data described above prior to analyses.**

I modified the data described above in three ways. First, I removed records that were not resolved to one day. Second, I removed records that were duplicated in terms of date, grid cell and species name. And finally, I reprojected records collected in Northern Ireland from the Irish national grid (OSNI 1951) to the British national grid (OSGB 1936).

```{r}
## process species occurrence data

# first remove data not resolved to one day

dat <- dat[- which(dat$startdate != dat$enddate)]

# then remove duplicates (in terms of species name, date and monad)

dat <- dat[- which(duplicated(dat[, c("recommended_name", "startdate", "monad")]))]

# drop columns that are not needed for analysis 

dat <- dat[, c("recommended_name", "monad", "startdate")]

# extract coordinates from grid references (needed by occAssess)

coords <- BRCmap::gr_let2num(gridref = dat$monad,
                             centre = TRUE,
                             return_projection = TRUE)

dat <- cbind(dat, coords)

# check if there are any coordinates on the OSNI projection

table(coords$PROJECTION)

# if yes then reproject these onto OSGB

if ("OSNI" %in% coords$PROJECTION) {
  
  GBCRS <- sp::CRS("+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +datum=OSGB36 +units=m +no_defs")
  
  NICRS <- sp::CRS("+proj=tmerc +lat_0=53.5 +lon_0=-8 +k=1 +x_0=200000 +y_0=250000 +ellps=airy +towgs84=482.5,-130.6,564.6,-1.042,-0.214,-0.631,8.15 +units=m +no_defs")
  
  datNI <- dat[which(dat$PROJECTION == "OSNI"), ]
  
  datGB <- dat[which(dat$PROJECTION == "OSGB"), ]

  NIcoords <- datNI[, c("EASTING", "NORTHING")]

  sp::coordinates(NIcoords) <- c("EASTING", "NORTHING")

  sp::proj4string(NIcoords) <- NICRS

  NIcoords <- sp::spTransform(NIcoords, GBCRS)

  datNI[,c("EASTING", "NORTHING")] <- data.frame(NIcoords)
  
  dat <- rbind(datGB, datNI)
  
}

# remove more columns that aren't needed

dat <- dat[, c("recommended_name", "startdate", "EASTING", "NORTHING")]

head(dat)

# create a new column for year (needed by occAssess). Note we'll keep date as it will allow
# us to look specifically at repeat visits later 

dat$year <- substr(dat$startdate, 1, 4)

# create identifier and sptialUncertainty fields (again, needed by occAssess)

dat$identifier <- "all_data"

dat$spatialUncertainty <- 1000

head(dat)

## now create a second dataset with just the repeat visits 

repeats <- dat[which(duplicated(dat[, c("EASTING", "NORTHING", "startdate")])), ]

repeats$identifier <- "repeat_visits" # set identifier to distinguish from the rest

# append to dat for analysis with occAssess

dat <- rbind(dat, repeats)
 
```

# 3.	Bias assessment and mitigation

## Assessment resolutions

**3.1 At what geographic, temporal and taxonomic resolutions (i.e. scales or grain sizes) will you conduct your bias assessment?**

I conducted the bias assessment at spatial and temporal resolutions of 1km and one year to match the statistical population about which I want to draw inferences (Table 2). However, it was not possible to assess the data at the species level; presence-only data say nothing about the spatial and temporal distribution of sampling where the focal species was not observed. Rather, I used the target group approach (Phillips et al., 2009) to approximate sampling effort, which is to say, I treated the spatial and temporal distribution of records for the whole taxonomic group (target group) as a proxy for the spatial and temporal distributions of sampling effort. In other words, if at least one species was recorded in some grid cell and at some time, then I assume that all species were searched for. 

## Geographic domain

**3.2 Are the data sampled from a representative portion of geographical space in the domain of interest?** 

<insert text>

**3.3 Are your data sampled from the same portions of geographic space across time periods?**

<insert text>

**3.4 If the answers to the above questions revealed any potential geographic biases, or temporal variation in geographic coverage, please explain, in detail, how you plan to mitigate them.** 

<insert text>

## Environmental domain

**3.5 Are your data sampled from a representative portion of environmental space in the domain of interest?** 

<insert text>

**3.6 Are your data sampled from the same portion of environmental space across time periods?**

<insert text>

**3.7 If the answers to the above questions revealed any potential environmental biases, or temporal variation in environmental coverage, please explain, in detail, how you plan to mitigate them.** 

<insert text>

## Taxonomic domain (or other organismal domain, e.g., phylogenetic, trait space etc.)

**Is the sampled portion of the taxonomic (or phylogenetic, trait or other space if more relevant) space representative of the taxonomic (or other) domain of interest?**

<insert text>

**3.9 Do your data pertain to the same taxa/taxonomic domain across time periods?** 

<insert text>

**3.10 If the answers to the above questions revealed any potential taxonomic biases, or temporal variation in taxonomic coverage, please explain, in detail, how you plan to mitigate them.** 

<insert text>

## Other potential biases

**3.11 Are there other potential temporal biases in your data that relate to variables other than ecological states?**

<insert text>

**3.12 Are you aware of any other potential biases not covered by the above questions that might cause problems for your inferences?**

<insert text>

**3.13 If questions 3.11 or 3.12 revealed any important potential biases, please explain how you will mitigate them.**

<insert text>

# 4.	Supporting references

<insert text>

